Supplementary Materials

### 2022-07-08 supplement/supplement-main.Rmd
paper revisions

---
title: "Supporting Information"
subtitle: "Modeling and leveraging intuitive theories to improve vaccine attitudes"
author: "Derek Powell, Kara Weisman, and Ellen M. Markman"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
bibliography: references.bib
csl: apa7.csl
figsintext : yes
always_allow_html: true
header-includes:
 \usepackage{float}
---

@@

1. When two beliefs are related by an intuitive theory, evidence affecting one of those beliefs will also affect the other in accordance with their relationship in the intuitive theory (e.g. evidence for B will increase credence in A, and vice versa). This is a clear and direct consequence of the laws of probability and Bayes' rule.
2. Where beliefs about two states of affairs, A and B, are related through an intuitive theory (e.g., A causes B), those beliefs will be systematically correlated across individuals in proportion to the average conditional probabilities (e.g. people who believe A will also tend to believe B);

@@

$p_i(y) = p_i(y|x)p_i(x) + p_i(y|\neg x)(1-p_i(x))$

@@

$p_i(y) = p_i(y|\neg x) + (p_i(y|x)p_i(x) - p_i(y|\neg x))p_i(x)$

@@

3. Thus we can often reasonably expect there will be some systematic relationship between the average change in each belief following evidence and the correlation among those beliefs across individuals. More specifically, if we assume that the changes in beliefs A following evidence are uncorrelated with the conditional probabilties connecting A and B in people's intuitive theories, then the average change in belief A and B following evidence affecting A will be proportional to the correlation of those beliefs across individuals.

@@

First, we expand $\Delta p_i(y)$ via the law of total probability and rearrange terms
$$
\begin{align*}
\Delta p_i(y)  &= p_i(y|e) - p_i(y)\\
 &= p_i(y|x)p_i(x|e) + p_i(y|\neg x)(1-p_i(x|e)) - p_i(y|x)p_i(x) + p_i(y|\neg x)(1-p_i(x)) \\
&= \big(p_i(x|e) - p_i(x)\big)\cdot \big(p_i(y|x) - p_i(y|\neg x)\big)
\end{align*}
$$
Taking the average, we have
$$
\overline{\Delta p_i(y)} = \overline{p(y|e) - p(y)} = \frac{1}{n}\sum_i\big(p_i(x|e) - p_i(x)\big)\big(p_i(y|x) - p_i(y|\neg x)\big)
$$
From the definition of covariance we have
$$
cov(X,Y) = E(XY) - E(X)E(Y)
$$
If we assume that the change in belief in x following the evidence, $\Delta p_i(x)$ is uncorrelated with the conditional probabilities encoded in the intuitive theory $p_i(y|x)p_i(x) - p_i(y|\neg x)$ (i.e. covariance equals zero) and apply this definition, we can then rewrite the average change as
$$
\overline{\Delta p_i(y)} = \overline{p(y|e) - p(y)} = \frac{1}{n}\sum_ip_i(x|e) - p_i(x) \frac{1}{n}\sum_ip_i(y|x) - p_i(y|\neg x)
$$
And we have thus shown that $\overline{\Delta p_i(y)} \propto \overline{p_i(y|x)p_i(x) - p_i(y|\neg x)} \propto corr(p_i(x),p_i(y))$

In practice we did not measure probabilistic credences directly, but rather asked participants to report on a agree/disagree scale. We assume these actual measurements are some monotonic function of these credences, $f(p_i(x))$. Allowing that $f$ could be non-linear, the proportionality conclusions are no longer guaranteed.

@@

```{r fig1-dag-example, echo=F, message=F, warning=F, fig.align='center', out.width = '40%', fig.cap="Example DAG illustrating some variables related to wildfires."}

library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

grViz(
  diagram = "
  digraph causal {

	  # Nodes
	  node [shape = oval]
	  D [label = 'Dry']
	  H [label = 'Heat']
	  F [label = 'Fire']
	  S [label = 'Smoke']

	  # Edges
	  edge [color = black,
	        arrowhead = vee]
	  rankdir = LR
	  D->F
	  H->F
	  F->S

	  # Graph
	  graph [overlap = true, fontsize = 8]
	}"
)
```

@@

$$P(y) = \Big(1-(1-w_0)\prod_i(1-w_i)^{x_i}\Big)\prod_j(1-w_j)^{x_j}$$

@@

$$y \sim beta\big(\mu k, (1-\mu)k\big)$$
$$\mu = \exp\Bigg(\ln\Big(1 - \exp\big(\ln\big(1-w_0) + \sum_i(1-w_i)x_i\big)\big) \Big) + \ln\Big(\sum_i(1-w_i)(1-x_i)\Big)\Bigg)$$

@@

Fortunately, the formalism of Bayesian Networks provides tools for navigating these challenges. Here, we used Bayesian network structure learning algorithms to construct a model of the intuitive theory of vaccine decisions (see Scutari & Denis, 2021). One way to learn a Bayesian network structure from data is to search through a space of networks, “score” each and choose the best scoring network. Many scoring metrics are possible, but a directly meaningful one is p(d|G), the likelihood of the data given the graph structure. Asymptotically, this can be estimated using the Bayesian Information Criterion (BIC) (citation):

$$BIC = k \ln (n) - 2\sum_{i}\hat{\ell}(x_i|x_{va(x_i)})$$

@@

library(tidyverse)
library(brms)
library(bnlearn)
library(HydeNet)
library(parallel)

@@

$$ log\big(O(h|d)\big) = \ln\big(O(h)\big) + \ln\Big(\frac{P(d|h=1)}{P(d|h=0)}\Big) $$

@@

$$ER = \frac{P(d|h=1)}{P(d|h=0)}$$

@@

$$y \sim beta\big(\mu k, (1-\mu)k\big)$$
$$logit(\mu) = logit(pretest) + \ln(ER_0) + x\ln(ER_{1})$$

@@

$$ p(evidence = 1 | x ) = .50*ER_1$$
$$ p(evidence = 0 | x) = .50. $$

@@

```{r}
## Effect size for study 3
s3_reg %>% 
  group_by(workerId, condition, phase) %>% 
  summarize(mresp = mean(response, na.rm=TRUE)) %>% 
  spread(phase, mresp) %>% 
  ungroup() %>% 
  group_by(condition) %>% 
  mutate(
    sd_pre = sd(pre),
    diff = post - pre
  ) %>% 
  summarize(
    m_diff = mean(diff),
    sd_pre = mean(sd_pre)
  ) %>% 
  mutate(
    cohen_d = m_diff/sd_pre
  )

>>>>

---
title: "Supplementary Materials"
subtitle: "Modeling and leveraging intuitive theories to improve vaccine attitudes"
author: "Derek Powell, Kara Weisman, and Ellen M. Markman"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
bibliography: references.bib
csl: apa7.csl
figsintext : yes
always_allow_html: true
header-includes:
 \usepackage{float}
---

@@

1\. When two beliefs are related by an intuitive theory, evidence affecting one of those beliefs will also affect the other in accordance with their relationship in the intuitive theory (e.g. evidence for B will increase credence in A, and vice versa). This is a clear and direct consequence of the laws of probability and Bayes' rule.

2\. Where beliefs about two states of affairs, A and B, are related through an intuitive theory (e.g., A causes B), those beliefs will be systematically correlated across individuals in proportion to the average conditional probabilities (e.g. people who believe A will also tend to believe B);

@@

\begin{equation}
p_i(y) = p_i(y|x)p_i(x) + p_i(y|\neg x)(1-p_i(x))
\end{equation}

@@

\begin{equation}
p_i(y) = p_i(y|\neg x) + (p_i(y|x)p_i(x) - p_i(y|\neg x))p_i(x)
\end{equation}

@@

3\. Thus we can often reasonably expect there will be some systematic relationship between the average change in each belief following evidence and the correlation among those beliefs across individuals. More specifically, if we assume that the changes in beliefs A following evidence are uncorrelated with the conditional probabilties connecting A and B in people's intuitive theories, then the average change in belief A and B following evidence affecting A will be proportional to the correlation of those beliefs across individuals.

@@

First, we expand $\Delta p_i(y)$ via the law of total probability and rearrange terms

\begin{align}
\Delta p_i(y)  &= p_i(y|e) - p_i(y)\\
 &= p_i(y|x)p_i(x|e) + p_i(y|\neg x)(1-p_i(x|e)) - p_i(y|x)p_i(x) + p_i(y|\neg x)(1-p_i(x)) \\
&= \big(p_i(x|e) - p_i(x)\big)\cdot \big(p_i(y|x) - p_i(y|\neg x)\big)
\end{align}

Taking the average, we have

\begin{equation}
\overline{\Delta p_i(y)} = \overline{p(y|e) - p(y)} = \frac{1}{n}\sum_i\big(p_i(x|e) - p_i(x)\big)\big(p_i(y|x) - p_i(y|\neg x)\big)
\end{equation}

From the definition of covariance we have

\begin{equation}
cov(X,Y) = E(XY) - E(X)E(Y)
\end{equation}

If we assume that the change in belief in x following the evidence, $\Delta p_i(x)$ is uncorrelated with the conditional probabilities encoded in the intuitive theory $p_i(y|x)p_i(x) - p_i(y|\neg x)$ (i.e. covariance equals zero) and apply this definition, we can then rewrite the average change as

\begin{equation}
\overline{\Delta p_i(y)} = \overline{p(y|e) - p(y)} = \frac{1}{n}\sum_ip_i(x|e) - p_i(x) \frac{1}{n}\sum_ip_i(y|x) - p_i(y|\neg x)
\end{equation}

And we have thus shown that $\overline{\Delta p_i(y)} \propto \overline{p_i(y|x)p_i(x) - p_i(y|\neg x)} \propto corr(p_i(x),p_i(y))$

In practice we did not measure probabilistic credences directly, but rather asked participants to report on a agree/disagree scale. We assume these actual measurements are some monotonic function of these credences, $f(p_i(x))$. Allowing that $f$ could be non-linear, the proportionality conclusions are no longer guaranteed, though a monotonic association would remain.

@@

```{r, out.width='100%', fig.align='center', fig.cap = 'Several examples illustrating different features of Bayesian Networks. \\newline A) A graphical representation of influence relations: Fire causes smoke. This influence is represented by the conditional dependencies captured in a conditional probability table (CPT). \\newline B) A “parent” node can be linked to multiple “children” and a “child” node can be linked to multiple “parents”: A child and its parents together compose a “family”. This graph has three families: {Accidents, lightning, drought, fire}, {fire, smoke}, and {fire, ash} described by a corresponding CPT (not pictured). \\newline C) Not all influence relations correspond to causation: A storm consists of rain, thunder, and lightning (marked as dashed arrows). Causes can be generative or preventive, e.g. rain prevents drought and fire (red arrows). Some nodes may exert countervailing influences on others: A storm consists of rain, which can prevent fires, but a storm also consists of lightning, which can cause fires. In addition, parents may influence their children independently or interactively. Lightning and accidents could be seen as independent causes of fire—an accident may cause a fire whether or not there is lightning, and vice versa. However, drought conditions interact with both of these other parent nodes, making them more likely to produce a fire. In the present work, we will consider only the role of independent influence relations.'}
knitr::include_graphics("../paper/fig-bn-examples.png")
```

@@

\begin{equation}
P(y) = \Big(1-(1-w_0)\prod_i(1-w_i)^{x_i}\Big)\prod_j(1-w_j)^{x_j}
\end{equation}

@@

\begin{align}
y &\sim beta\big(\mu k, (1-\mu)k\big) \\
\mu &= \exp\Bigg(\ln\Big(1 - \exp\big(\ln\big(1-w_0) + \sum_i(1-w_i)x_i\big)\big) \Big) + \ln\Big(\sum_i(1-w_i)(1-x_i)\Big)\Bigg)
\end{align}

@@

Fortunately, the formalism of Bayesian Networks provides tools for navigating these challenges. Here, we used Bayesian network structure learning algorithms to construct a model of the intuitive theory of vaccine decisions (see Scutari & Denis, 2021). One way to learn a Bayesian network structure from data is to search through a space of networks, “score” each and choose the best scoring network. Many scoring metrics are possible, but a directly meaningful one is p(d|G), the likelihood of the data given the graph structure. Asymptotically, this can be estimated using the Bayesian Information Criterion (BIC):

\begin{equation}
BIC = k \ln (n) - 2\sum_{i}\hat{\ell}(x_i|x_{va(x_i)})
\end{equation}

@@

library(tidyverse)
library(brms)
library(bnlearn)
library(HydeNet)
library(parallel)
library(DiagrammeR)

@@

\begin{equation}
log\big(O(h|d)\big) = \ln\big(O(h)\big) + \ln\Big(\frac{P(d|h=1)}{P(d|h=0)}\Big) 
\end{equation}

@@

\begin{equation}
ER = \frac{P(d|h=1)}{P(d|h=0)}
\end{equation}

@@

\begin{align}
y &\sim beta\big(\mu k, (1-\mu)k\big) \\
logit(\mu) &= logit(pretest) + \ln(ER_0) + x\ln(ER_{1})
\end{align}

@@

\begin{align}
p(evidence = 1 | x ) &= .50*ER_1 \\
p(evidence = 0 | x) &= .50. 
\end{align}

@@

```{r, include=F}
## Effect size for study 3
s3_reg %>% 
  group_by(workerId, condition, phase) %>% 
  summarize(mresp = mean(response, na.rm=TRUE)) %>% 
  spread(phase, mresp) %>% 
  ungroup() %>% 
  group_by(condition) %>% 
  mutate(
    sd_pre = sd(pre),
    diff = post - pre
  ) %>% 
  summarize(
    m_diff = mean(diff),
    sd_pre = mean(sd_pre)
  ) %>% 
  mutate(
    cohen_d = m_diff/sd_pre
  )
